High-Dimensional Regularized Discriminant Analysis

### 2013-12-12 reports/paper-hdrda.Rnw
Began rewrite of abstract and intro to focus on properties of RDA and its reparameterization. Related to #10 and #15.

\title{Efficient Calculation of Regularized Discriminant Analysis for
  High-Dimensional Classification}

\author{John A. Ramey, Caleb K. Stein, \and Dean M. Young}

@@

Reliable supervised classification of tumors from gene expression microarray
data is essential for successful diagnosis and treatment of diseases such as
cancer; however, because the number of gene expression levels measured is
typically much greater than the sample size, the analysis of microarray data
often presents difficult statistical and computational challenges. Here, we
propose a new regularized discriminant analysis classifier that incorporates
dimension reduction and covariance-matrix shrinkage to enable a computationally
efficient classifier. We demonstrate with four high-dimensional microarray data
sets that our proposed classifier is often superior to several recently proposed
sparse and regularized classifiers in terms of classification accuracy. Finally,
we provide an implementation of our proposed classifier in the {\tt
  sparsediscrim} R package, available on CRAN.

@@

Microarrays have enabled researchers to measure gene-expression levels of
thousands of genes in a single experiment to identify a subset of genes and
their expression patterns that either characterize a particular cell state or
facilitate an accurate prediction of future cell state.  We focus on the
classification of tissues and other biological samples into one of $K$ disjoint,
predefined types, where our goal is to ascertain the class membership of
a new tissue sample based on gene-expression microarray data. However,
microarray data typically consist of a large number of gene expressions $p$
relative to the total sample size $N$. Moreover, standard benchmark microarray
data sets usually consist of between $p = 10,000$ and $55,000$ probes and fewer
than $N = 100$ observations, in which case standard supervised-learning methods
exhibit poor classification performance because of the \emph{curse of
  dimensionality} \citep{Bellman:1961tn}. Furthermore, some well-known
classifiers, such as linear and quadratic discriminant analysis, are
incalculable when $p > N$ because the sample covariance matrices employed are
singular. As a result, many researchers, including \cite{Mai:2012bf},
\cite{Fan:2012iq}, \cite{Merchante:2012vk}, \cite{Clemmensen:2011kr}, and
\cite{Huang:2010ju}, have introduced restrictions on the population covariance
matrices in linear and quadratic discriminant analysis to reduce the number of
covariance parameters that must be estimated to ensure that the resulting
classifiers are calculable. For instance, \cite{Dudoit:2002ev} have modified the
linear and quadratic discriminant analysis classifiers by assuming that the
population covariance matrices are diagonal, resulting in classifiers that are
special cases of the Naive Bayes classifier \citep[Chapter 4]{Hastie:2008dt}.

Alternatively, several researchers, including \cite*{Zhang:2010va},
\cite{Ye:2009gd}, \cite{Ji:2008wp}, \cite*{Guo:2007te},
\cite*{Srivastava:2007ud}, \cite{Ye:2006vx}, and \cite{Ye:2005uu}, have proposed
regularization methods to improve the estimation of the population covariance
matrices before applying either linear or quadratic discriminant
analysis. Covariance-matrix regularization methods often are based on the
well-known ridge-regression approach of \cite{Hoerl:1970cd} and adjust the
eigenvalues of the sample covariance matrices to ensure positive
definiteness. For excellent overviews of the regularization problem in
discriminant analysis, see \cite{Murphy:2012uq}, \cite{Mkhadri:1997gy}, and
\cite{Friedman:1989tm}.

@@

In discriminant analysis we wish to assign correctly an unlabeled
$p$-dimensional observation vector $\bm x$ to one of $K$ unique, known classes
(or populations) by constructing a classifier from $N$ training
observations. Let $\bm x_i = (x_{i1}, \ldots, x_{ip}) \in \mathbb{R}_{p \times
  1}$ be the $i$th observation $(i = 1, \ldots, N)$ with true, unique membership
$y_i \in \{\omega_1, \ldots, \omega_K\}$. We assume that $(\bm x_i, y_i)$ is a
realization from a mixture distribution $p(\bm x) = \sum_{k=1}^K p(\bm x |
\omega_k) p(\omega_k)$, where $p(\bm x | \omega_k)$ is the probability density
function of the $k$th class and $p(\omega_k)$ is the prior probability of class
membership of the $k$th class. We further assume that each of $p(\omega_k)$ is
equal, $k = 1, \ldots, K$.

@@

@@

>>>>

\title{High-Dimensional Regularized Discriminant Analysis}

\author{John A. Ramey, Caleb K. Stein, and Dean M. Young}

@@

LDA and QDA suck. Friedman to the rescue with RDA. We reformulate the RDA
classifier. Our parameterizaton yields 1) improved interpretation, 2) nice
properties that facilitate dimension reduction, and 3) as a result of the
dimension reduction, efficient calculation for high-dimensional data.

Here, we propose a new regularized discriminant analysis classifier that
incorporates dimension reduction and covariance-matrix shrinkage to enable a
computationally efficient classifier. We demonstrate with four high-dimensional
microarray data sets that our proposed classifier is often superior to several
recently proposed sparse and regularized classifiers in terms of classification
accuracy. Finally, we provide an implementation of our proposed classifier in
the {\tt sparsediscrim} R package, available on CRAN.

@@

Linear discriminant analysis (LDA) and quadratic discriminant analysis (QDA) are
classical approaches to supervised learning and are widely used in a variety of
areas, such as medical research, facial recognition, and TODO
\cite{Dudoit:2002ev,TODO1,TODO2} Both the LDA and QDA classifiers are based on
the assumption that the underlying distribution of each of the $K$ classes are
multivariate normally distributed. While both methods are popular in practice,
they often cannot be applied to modern data sets because the sample covariance
matrices employed are singular when the number of features exceeds the total
sample size $N$. For instance, standard benchmark microarray data sets used in
biological research often consist of between $p = 10,000$ and $55,000$ probes
and fewer than $N = 100$ observations, in which case LDA and QDA are
incalculable.

\cite{Friedman:1989tm} proposed regularized discriminant analysis (RDA), which
partially pools the individual sample covariance matrices used in the QDA
classifier with the pooled sample covariance matrix from the LDA
classifier. After pooling the covariance matrices, the RDA classifier employs a
covariance-matrix regularization technique that shrinks the zero eigenvalues of
the resulting covariance matrices to ensure they are nonsingular. Similarly,
several researchers, including \cite*{Zhang:2010va}, \cite{Ye:2009gd},
\cite{Ji:2008wp}, \cite*{Guo:2007te}, \cite*{Srivastava:2007ud},
\cite{Ye:2006vx}, and \cite{Ye:2005uu}, have proposed regularization methods to
improve the estimation of the population covariance matrices before applying
either the LDA or QDA classifiers. Covariance-matrix regularization methods
often are based on the well-known ridge-regression approach of
\cite{Hoerl:1970cd} and adjust the eigenvalues of the sample covariance matrices
to ensure positive definiteness. For excellent overviews of the regularization
problem in discriminant analysis, see \cite{Murphy:2012uq},
\cite{Mkhadri:1997gy}, and \cite{Friedman:1989tm}.

% TODO: Should we remove this paragraph? Keep a couple of the recent references
% though.
Alternatively, many researchers, including \cite{Mai:2012bf}, \cite{Fan:2012iq},
\cite{Merchante:2012vk}, \cite{Clemmensen:2011kr}, and \cite{Huang:2010ju}, have
introduced restrictions on the population covariance matrices in linear and
quadratic discriminant analysis to reduce the number of covariance parameters
that must be estimated to ensure that the resulting classifiers are
calculable. For instance, \cite{Dudoit:2002ev} have modified the linear and
quadratic discriminant analysis classifiers by assuming that the population
covariance matrices are diagonal, resulting in classifiers that are special
cases of the Naive Bayes classifier \citep[Chapter 4]{Hastie:2008dt}.

We reformulate the RDA classifier. Our parameterizaton yields 1) improved
interpretation, 2) nice properties that facilitate dimension reduction, and 3)
as a result of the dimension reduction, efficient calculation for
high-dimensional data.

@@

In discriminant analysis we wish to assign correctly an unlabeled vector $\bm x
\in \mathbb{R}_{p \times 1}$ to one of $K$ unique, known classes by constructing
a classifier from $N$ training observations. Let $\bm x_i = (x_{i1}, \ldots,
x_{ip}) \in \mathbb{R}_{p \times 1}$ be the $i$th observation $(i = 1, \ldots,
N)$ with true, unique membership $y_i \in \{\omega_1, \ldots, \omega_K\}$. We
assume that $(\bm x_i, y_i)$ is a realization from a mixture distribution $p(\bm
x) = \sum_{k=1}^K p(\bm x | \omega_k) p(\omega_k)$, where $p(\bm x | \omega_k)$
is the probability density function of the $k$th class and $p(\omega_k)$ is the
prior probability of class membership of the $k$th class. We further assume that
each of $p(\omega_k)$ is equal, $k = 1, \ldots, K$.

@@

\cite{Halbe:2007ea} demonstrate that  parameterization results in a

@@

\subsection{Relationship with Regularized Discriminant Analysis}

\cite{Friedman:1989tm} proposed regularized discriminant analysis (RDA) by
incorporating a weighted average of the sample covariance matrix $\widehat{ \bm
  \Sigma}_k$ and the pooled sample covariance matrix $\widehat{\bm \Sigma}$ to
estimate $\bm \Sigma_k$ with
\begin{align}
    \widehat{\bm \Sigma}_k(\lambda) = n_k^{-1}(\lambda) \bm S_k(\lambda) \quad
    (k =
      1, \ldots, K),\label{eq:sig-lambda}
      \end{align}
where $\lambda \in [0, 1]$, $n_k(\lambda) = (1 - \lambda) n_k + \lambda N$, $
\bm S_k = n_k \widehat{\bm \Sigma}_k$, $\bm S = \sum_{k=1}^K \bm S_k$, and $ \bm
S_k(\lambda) = (1 - \lambda) \bm S_k + \lambda \bm S$. If $\lambda = 0$,
\eqref{eq:sig-lambda} is equal to $\widehat{\bm \Sigma}_k$ used in
\eqref{eq:qda}. Alternatively, if $\lambda = 1$, \eqref{eq:sig-lambda} reduces
to \eqref{eq:pooled-cov}.

To further improve the estimation of $\bm \Sigma_k$ and to stabilize the inverse
of \eqref{eq:sig-lambda}, \cite{Friedman:1989tm} has proposed the biased
covariance-matrix estimator
\begin{align}
          \widehat{\bm \Sigma}_k(\lambda, \gamma) = (1 - \gamma) \widehat{\bm
            \Sigma}_k(\lambda) + \gamma \frac{\tr\{\widehat{\bm
              \Sigma}_k(\lambda)\}}{p} \bm I_p,\label{eq:sig-rda}
          \end{align}
where $\gamma \in [0, 1]$ is a regularization parameter that controls the
shrinkage of \eqref{eq:sig-rda} towards $\bm I_p$ and is weighted by the average
of the eigenvalues of \eqref{eq:sig-lambda}. Finally, the RDA classifier is
established by substituting \eqref{eq:sig-rda} into \eqref{eq:qda}.

\section{Properties of the HDRDA}
